Hidden Multiple Comparisons Increase Forensic Error Rates

### 2024-04-02 PNAS/index-old.tex
Removing coloration from index-old and remaking the diff file

A forensics examiner tasked with determining whether a wire in evidence was cut by a specific tool will create one or more blade cuts using the putative tool, which are then compared to the cut surface of the wire.
These cuts are made in a sheet of material similar to the wire's composition, and may be performed at multiple angles, as the angle of the tool to the substrate can affect which striations are recorded on the substrate surface.
The blade cuts will then be compared to the wire under a comparison microscope, though eventually, automatic comparison algorithms may also be validated for lab use.
Each side of each blade cut will be compared to each side of the wire;
\svp{different tool designs have between 2 and 4 cutting surfaces in contact with the substrate.}
% depending on the tool design, there can be between two and four cutting surfaces in contact with the substrate.

@@

\section*{Probability of False Identifications}
There are at least two components of the false positive rate (FPR): identifying two pieces of evidence that have similar characteristics but are from different sources (a coincidental match) and procedural failures (e.g.\ lab process errors) \citep[p 50]{pcast}.
In objective disciplines with standardized evaluation rules (e.g.\ DNA), these sources can be distinguished.
However, in toolmark examination, no objective evaluation rules are used; examiners testify based on subjective, individual rules for how much similarity is sufficient for an identification.
Assuming that lab procedure errors are not a factor in studies, we use reported error rates from open-set studies of striated evidence comparisons to estimate the coincidental match rate of a single wire-cut comparison.
The three studies we consider \citep{bajic2020, mattijssen2021,best2022} have FPRs between 0.0045 \citep{best2022} and 0.072 \citep{mattijssen2021}; pooling data from these studies yields an FPR of 0.02.
For a single-comparison FPR of $e$, the family-wise FPR for $n$ comparisons, $E_n$ is $1 - [1-e]^n$.
\Cref{tab:family-wise} \hh{shows the impact the number of comparisons has on these published error rates. With an error rate of 0.07, as suggested by Bajic (2020), examiners can make up to 14 comparisons, i.e.} \svp{even the simple example in this paper exceeds} \hh{an upper bound of 10\% for the family wise false identification error.}
%\Cref{tab:Accuracy} shows the family-wise accuracy rate on the y-axis; if 90\% is an acceptable overall accuracy, 2 comparisons can be made when $e = 0.05$; increasing to 10 comparisons when $e =  0.01$.
%Thus, at an error rate of sprintf("e = %.2f", fpr), examiners can conduct comparisons before exceeding the family-wise FPR of 10\%.
\svp{To conduct a search of a modestly sized database with 1000 entries, the initial FPR cannot}
% \hh{For a modestly sized database search of size 1000, the initial false positive rate should not
\hh{exceed 1 in 10,000 to guarantee a family-wise total false identification error of at most 10\%. }

Under these constraints, the accuracy of an examination involving multiple comparisons between a wire and a tool will be low, as the number of candidate alignments that must be examined is high.
Even the most innocuous example (small blade, only 2 cutting surfaces, and a relatively large wire) involves a minimum of 15 %completely independent
comparisons.
Examiners would make cuts under multiple angles \citep{baikerToolmarkVariabilityQuality2015}, increasing the number of comparisons \svp{and making a} false identification even more \svp{probable}.
As a result, it is questionable whether wire comparisons made under current protocols are reliable enough to be presented at trial.

\svp{Clearly, studies for wire evidence, and larger studies for striated evidence in general, are necessary.}
% \hh{Obviously, we need studies for wire evidence. It is also clear, based on the error rates, that larger studies are needed. }
\svp{Moving away from binary assessments toward quantification of striation similarity and observed pattern frequency will also reduce the severity of this issue and allow examiners to assign unusual striation patterns more weight in the process.}
% \hh{Going forward, we need to move away from  the binary assessment of 'match' or 'non-match' and quantify (a) the similarity of striations expressed between two pieces of evidence, and (b) the frequency of the observed pattern. Unusual striation patterns should be assigned more weight in the process.  }

>>>>

A forensics examiner tasked with determining whether a wire in evidence was cut by a specific tool will create one or more blade cuts using the putative tool, which are then compared to the cut surface of the wire.
These cuts are made in a sheet of material similar to the wire's composition, and may be performed at multiple angles, as the angle of the tool to the substrate can affect which striations are recorded on the substrate surface.
The blade cuts will then be compared to the wire under a comparison microscope, though eventually, automatic comparison algorithms may also be validated for lab use.
Each side of each blade cut will be compared to each side of the wire;
different tool designs have between 2 and 4 cutting surfaces in contact with the substrate.

@@

\section*{Probability of False Identifications}
There are at least two components of the false positive rate (FPR): identifying two pieces of evidence that have similar characteristics but are from different sources (a coincidental match) and procedural failures (e.g.\ lab process errors) \citep[p 50]{pcast}.
In objective disciplines with standardized evaluation rules (e.g.\ DNA), these sources can be distinguished.
However, in toolmark examination, no objective evaluation rules are used; examiners testify based on subjective, individual rules for how much similarity is sufficient for an identification.
Assuming that lab procedure errors are not a factor in studies, we use reported error rates from open-set studies of striated evidence comparisons to estimate the coincidental match rate of a single wire-cut comparison.
The three studies we consider \citep{bajic2020, mattijssen2021,best2022} have FPRs between 0.0045 \citep{best2022} and 0.072 \citep{mattijssen2021}; pooling data from these studies yields an FPR of 0.02.
For a single-comparison FPR of $e$, the family-wise FPR for $n$ comparisons, $E_n$ is $1 - [1-e]^n$.
\Cref{tab:family-wise} shows the impact the number of comparisons has on these published error rates. With an error rate of 0.07, as suggested by Bajic (2020), examiners can make up to 14 comparisons, i.e. even the simple example in this paper exceeds an upper bound of 10\% for the family wise false identification error.
To conduct a search of a modestly sized database with 1000 entries, the initial FPR cannot
exceed 1 in 10,000 to guarantee a family-wise total false identification error of at most 10\%.

Under these constraints, the accuracy of an examination involving multiple comparisons between a wire and a tool will be low, as the number of candidate alignments that must be examined is high.
Even the most innocuous example (small blade, only 2 cutting surfaces, and a relatively large wire) involves a minimum of 15 %completely independent
comparisons.
Examiners would make cuts under multiple angles \citep{baikerToolmarkVariabilityQuality2015}, increasing the number of comparisons and making a false identification even more probable.
As a result, it is questionable whether wire comparisons made under current protocols are reliable enough to be presented at trial.

Clearly, studies for wire evidence, and larger studies for striated evidence in general, are necessary.
Moving away from binary assessments toward quantification of striation similarity and observed pattern frequency will also reduce the severity of this issue and allow examiners to assign unusual striation patterns more weight in the process.
